---
title: "Analysis of Editor Provenance"
output: html_document
---

# Summary

Our overall analyses of source publishers models the relationships between three entities:

 * The WP language edition.
 * The country associated with an article (ac).
 * The country associated with the editor of the article (ec).

We count the number of citations that occur for each possible triple, and find all possible attributes associated with ac, ec, and wplang significant.
Thus, we try to identify the most explanatory model variables.


*Localness:* 

For this analysis, ac = ec.
For example: Across all articles located in Japan (ac) in the Chinese Wikipedia (wplang), how many are edited by people Japan (ec).

Three country variables explain the vast majority of the overall variance:

  * If a country natively speaks the WP language edition, the oddsof a an editor being local increases by 17x (that's 1770%)
  * For every doubling in internet penetration, the odds an editor is local increase by 43%
  * For every increase of 10% in the amount a country represents of a language's worldwide base, the oddsof a publisher / editor being local increases by 41%. For example, for Japan (who represents almost all Japanese speakers) this would suggest an increase in odds of close to 410% ON TOP of the 1770% for language edition!
  
*Dipole provenance:*

For example: Across all articles located in Japan (ac) in the Chinese Wikipedia (wplang), how many are published by the U.S. (pc).

At a high-level we model the overall number of citations in a language edition that come from a country.
Our model accounts for the vast majority of the deviance, and identifies two variables as extremely important:

  * For every doubling in the number of journal articles in a country, the odds of a citation being from the country increase by 280%.
  * For every increase of 10% in the amount a country represents of a language's worldwide base, the oddsof a publisher / editor being local increases by 1100%!
  * The odds an edit comes from a country that natively speaks the WP language editon are 11X higher.
  
After we subtract off this "top-level" country effect, we look to model the relationship between an article's country (ac) and editor country (ec).
Three variables emerge as extremely significant:

  * For every doubling in the distance between countries, the odds of a country being associated an article in another country decrease by roughly 40%.
  * For every doubling in the number of journal articles for a country, the odds of a publisher in the country increase by roughly 50%
  * An editor from a country that natively speaks the wp lang is 390% more likely to edit an ac.

# Preparing the dataset

Import required modules for renaming and world bank data:

```{r}
options(width=120);
require('plyr');
require('WDI');
require('MASS');
```

Read in raw file, calculate totals for each editor country and "article" country.

```{r}
df <- read.table('../../../dat/editor-counts-enhanced.tsv', header=TRUE, sep="\t", fileEncoding="utf-8");
totals <- rename(aggregate(count ~ project + article_country, df, sum), c("count"="article_total"));
df <- merge(df, totals, by=c("project","article_country"));
totals <- rename(aggregate(count ~ project + other_country, df, sum), c("count"="other_total"));
df <- merge(df, totals, by=c("project","other_country"));
totals <- rename(aggregate(count ~ project, df, sum), c("count"="project_total"));
df <- merge(df, totals, by=c("project"));
```

Read in World bank data and fields: GDP, Population, tertiary grads, broadband population.
Rename these fields and merge them into our table and impute zeros for missing data (TODO: something better).

```{r}
fields <- c("NY.GDP.MKTP.CD", "SP.POP.TOTL", "SE.TER.GRAD", "IT.NET.BBND");
field_mapping <- c("GDP", "POP", "EDU", "INTERNET");
names(field_mapping) = fields;

wdi = WDI(indicator=fields, start=2012, end=2012);
wdi$iso2c <- tolower(wdi$iso2c);
wdi <- rename(wdi, field_mapping);
df <- merge(df, wdi, by.x="other_country", by.y="iso2c");

df$GDP[is.na(df$GDP)] <- 0;
df$POP[is.na(df$POP)] <- 0;
df$EDU[is.na(df$EDU)] <- 0;
df$INTERNET[is.na(df$INTERNET)] <- 0;
```
Create log-transformed versions of skewed fields:

```{r}
df$logmigration1 <- log(df$migration1 + 2);
df$logmigration2 <- log(df$migration2 + 2);
df$logjournals1 <- log(df$journals1 + 2);
df$logjournals2 <- log(df$journals2 + 2);
df$lognewspapers1 <- log(df$newspapers1 + 2);
df$lognewspapers2 <- log(df$newspapers2 + 2);
df$logcountrylangpop <- log(df$countrylangpop + 2);
df$logdist <- log(df$dist + 2);
df$logGDP <- log(df$GDP + 2);
df$logPOP <- log(df$POP + 2);
df$logEDU <- log(df$EDU + 2);
df$logINTERNET <- log(df$INTERNET + 2);
```

# Overview of dataset

Our dataset models the relationship between language edition and the dipole of article country (ac) and the country of the editor (ec).

Four fields contain the key information being aggregated together by this dataset:

 * project: WP language edition (wplang)
 * article_country: the country containing the article (ac)
 * other_country: the country associated with an editor (ec)
 * count: the number of edits or published citations associated with the triple (project, article_country, other_country) in our dataset.
 
 * article-is-native: whether the WP edition language is "natively" spoken by the ac
 * other-is-native: whether the WP edition language is "natively" spoken by the editor
 * distance: distance between ac and ec
 * migration1: the number of people who migrated from ac to ec in 2012 (UN data)
 * migration2: the reverse of migration1
 * newspapers1: the number of newspapers published in ac (UN data)
 * newspapers2: the number of newspapers published in ec (UN data)
 * journals1: the number of journal articles published in ac (from Scopus via SCImago)
 * journals2: the number of journal articles published in ec (from Scopus via SCImago)
 * wplangshare: the fraction of worldwide speakers of wplang that live in ec
 * countrylangshare: across all languages used in ac, the fraction of worldwide speakers that live in oc 
 * countrylangpop: across all languages used in ac, the number of people who speak the language in the ec
 * langsim: the cosine similarity between the languages spoken in ac and ec
 
 Note that we use "raw counts" throughout most attributes (not per-captia values).
 The justification behind this is that work output is more related to the raw number of people (and therefore potential editors / publishers).
 A country with 1000 people, but 90% broadband access is going to have a much smaller pool of total Wikipedia work than a country with 1M people and 10% broadband access.

```{r}
head(df);
```

### Correlation between features

Below you can see the correlation matrix between all the features.
Note that the basic demographic indicators (GDP, POP, etc) are all strongly correlated with each other (0.7 to 0.95). 
The other variables are largely uncorrelated with each other, with a few moderate exceptions (e.g. migration in both directions = 0.4, journal vs newspaper is 0.36, population and newspaper is 0.76, etc.)

Interestingly, although journals and GDP are extremely highly correlated (0.98), log transforming decreases this relationship significantly to 0.52. 
This is probably because GDP and journals show high correlation in extreme values (e.g. the US), but after log transformation these have a much smaller effect.

```{r}
cor(unique(df[,c("distance", "GDP", "POP", "EDU", "INTERNET", "migration1", "migration2", "journals1", "journals2", "newspapers1", "newspapers2", "countrylangpop", "wplangshare", "langsim", "countrylangshare")]));
```

# Localness

First, we consider a statistical model of localness. We begin by filtering the dataset down to those records that have the same article-country and other-country:

```{r}
local_df <- subset(df, as.character(article_country) == as.character(other_country));
head(local_df);
nrow(local_df);
```

We now build the full model. 
Since article_country and other_country are the same, we only include variables for one of the countries.
Note that all coefficients are significant!

```{r}
fit <- glm(cbind(count, article_total - count) ~ article_is_native + newspapers1 + journals1 + wplangshare + GDP + EDU + INTERNET + lognewspapers1 + logjournals1 + logGDP + logPOP + logEDU + logINTERNET, family = binomial(logit), data=local_df)
summary(fit);
```

Despite the significance of the model terms, it is helpful to prune the model down to a set of variables after which we see diminshing returns.
We use R's step function to consider every possible model, adding one variable at a time.

We begin by carefully analyzing the effects of all variables individually. 
article_is_native is the strongest independent signal of localness, explaining 58% of the model deviance, followed by wplangshare.

```{r}
min_fit <- glm(cbind(count, article_total-count) ~ 1, family = binomial(logit), data=local_df);
s <- step(min_fit, scope=formula(fit), direction='forward', steps=1);
```

We now build up the full model in stepwise fashion (starting with log journals).
Since log(journals) is highly correlated with logGDP (0.52), it unsurprisingly pushes GDP to the bottom of the equation.

```{r}
s <- step(min_fit, scope=formula(fit), direction='forward', trace='false');
s$anova;
```

We can now see that a few key variables can account for most of the model's explanatory power.
In particular, the first four variables: article_is_native, logINTERNET, wplangshare explain the vast majority of the model:


```{r}
small_fit <- glm(cbind(count, article_total - count) ~ article_is_native + logINTERNET + wplangshare, family = binomial(logit), data=local_df)
summary(small_fit);
```

```{r}

fit <- glm(count/article_total ~ article_is_native + newspapers1 + journals1 + wplangshare + GDP + EDU + INTERNET + lognewspapers1 + logjournals1 + logGDP + logPOP + logEDU + logINTERNET, family = binomial(logit), data=local_df)
summary(fit);
min_fit <- glm(count / article_total ~ 1, family = binomial(logit), data=local_df);
s <- step(min_fit, scope=formula(fit), direction='forward');
s$anova;
#small_fit <- glm(count/article_total  ~ article_is_native + logINTERNET + wplangshare, family = binomial(logit), data=local_df)
#summary(small_fit);
```

To interpret our model, we can look at the confidence intervals and odds ratios associated with it.

  * If a country natively speaks the WP language edition, the oddsof a an editor being local increases by 17x (that's 1770%)
  * For every doubling in internet penetration, the odds an editor is local increase by 43%
  * For every increase of 10% in the amount a country represents of a language's worldwide base, the oddsof a publisher / editor being local increases by 41%. For example, for Japan (who represents almost all Japanese speakers) this would suggest an increase in odds of close to 410% ON TOP of the 1770% for language edition!

```{r}
exp(cbind(oddsRatio=coef(small_fit), confint(small_fit)));
```



# Full dipole provenance

We now move on to modeling the dipole relationship.
For example, focusing on Japane, can how can we model how likely China is to be a publisher of a source?

### Modeling <language, country of source publisher>

We begin  by modeling top-level effects at the (language, source publisher country) level.
In other words, how likely is China to be the publisher of any citation in English Wikipedia?

We restrict our dataset to this top-level information:

```{r}
top_df <- unique(df[,c("project", "project_total", "other_total", "other_country", "other_is_native",  "newspapers2", "journals2", "wplangshare", "GDP", "POP", "EDU", "INTERNET", "lognewspapers2", "logjournals2", "logGDP", "logPOP", "logEDU", "logINTERNET")]);
head(top_df);
```

Now let's build and inspect a model.
Note that the "localness" problem appears to be much more difficult; this model explains 85% of the deviance in the dataset.

```{r}  
fit <- glm(cbind(other_total, project_total-other_total) ~ . - other_country - project - project_total - other_total, family = binomial(logit), data=top_df);
summary(fit);
```
Like before we will examine whether we can build a more compact and interpretable model.
We first look at the strength of each individual predictor.
Again, wplangshare rises to the top.

```{r}
min_fit <- glm(cbind(other_total, project_total-other_total) ~ 1, family = binomial(logit), data=top_df);
s <- step(min_fit, scope=formula(fit), direction='forward', steps=1);
```

We now run a forward stepwise regression. 
Three variables, wplangshare, log journals, other_is_native rise to the top.

```{r}
s <- step(min_fit, scope=formula(fit), direction='forward', trace=FALSE);
s$anova;
```

We inspect the pruend model:

```{r}
small_fit <- glm(cbind(other_total, project_total-other_total) ~ wplangshare + logjournals2 + other_is_native, family = binomial(logit), data=top_df);
summary(small_fit);
```

To interpret our model, we can look at the confidence intervals and odds ratios associated with it.

  * For every doubling in the number of journal articles in a country, the odds of a citation being from the country increase by 280%.
  * For every increase of 10% in the amount a country represents of a language's worldwide base, the oddsof a publisher / editor being local increases by 1100%!
  * The odds an edit comes from a country that natively speaks the WP language editon are 11X higher.

```{r}
exp(cbind(oddsRatio=coef(small_fit), confint(small_fit)));
```


### Modeling dipoles

Finally, we move on to dipoles. 
Since we have already accounted for top-level effects, we introduce a variable that models the fraction of all citations a country accounts for in each language edition.

```{r}
df$top_level <- df$other_total / df$project_total;
head(df);
```

We now try to model the remaining variance. 
Once again, all terms in the model are extremely significant and the full model accounts for the vast majority (92%) of the variance.

```{r}
full <- glm(cbind(count, article_total-count) ~ . - article_country - other_country - project -project_total -article_total - country, family = binomial(logit), data=df);
summary(full);
```

We now try to identify a smaller, more interpretable model.
Unsurprisingly, the background popularity (analyzed in the previous section) of a country as a source publisher is by far the strongest term.

We then move on to the second term in the model.
log(distance) between ac and ec is most predictive, followed by log(num-journals for ac)) and other-is-native.

```{r}
min_fit <- glm(cbind(count, article_total-count) ~ 1, family = binomial(logit), data=df);
s <- step(min_fit, scope=formula(full), direction='forward', steps=2);
s <- step(min_fit, scope=formula(full), direction='forward', steps=6, trace=FALSE);
s$anova;
```

```{r}
small_fit <- glm(cbind(count, article_total-count) ~ top_level + logdist + logjournals2 + other_is_native, family = binomial(logit), data=df);
summary(small_fit);
```

To interpret our model, we can look at the confidence intervals and odds ratios associated with it.

  * For every doubling in the distance between countries, the odds of a country being associated an article in another country decrease by roughly 40%.
  * For every doubling in the number of journal articles for a country, the odds of a publisher in the country increase by roughly 50%
  * An editor from a country that natively speaks the wp lang is 390% more likely to edit an ac.

```{r}
exp(cbind(oddsRatio=coef(small_fit), confint(small_fit)));
```
