---
title: "Analysis of Source Publisher Provenance"
output: html_document
---

# Summary

Our overall analyses of source publishers models the relationships between three entities:

 * wp-lang: The WP language edition.
 * ac: The country an article is about.
 * pc: The country where a source publisher is located.

We count the number of citations that occur for each possible triple of (wp-lang, ac, pc).


*Localness:* 

For this analysis, ac = pc.
For example: Across all articles about Japan (ac) in the Chinese Wikipedia (wplang), how many citations are published by Japanese publishers (pc).

Four country variables explain the vast majority of the overall variance:

* The number of journal of articles
* The country's population
* Whether a country speaks the same language as the language edition
* The share of speakers of the WP language edition a country accounts for explain the vast majority of the top-level contributions.

The effect sizes we see are quite large:

* For every doubling in the number of journal articles a country has published, the odds of its sources being local double.
* If a country natively speaks the language of the WP edition, the odds of a source being local to the country increase by 47%.

*Dipole provenance:*

For example: Across all articles located in Japan (ac) in the Chinese Wikipedia (wplang), how many are published by the U.S. (pc).

At a high-level we model the overall number of citations in a language edition that come from a country.
Our model accounts for 85% of the deviance, and identifies two variables as extremely important:

  * For every doubling in the number of journal articles in a country, the odds of a citation being from the country increase by 300%.
  * For every increase of 10% in the amount a country represents of a language's worldwide base, the odds a citation coming froms that country increase by 600%.
  
After we subtract off this "top-level" country effect, we look to model the relationship between an article's country (ac) and a publisher country (pc).
Again, two variables emerge as extremely significant:

  * For every doubling in the distance between ac and pc, the odds of pc publishing a source in ac are roughly halved.
  * For every doubling in the number of journal articles for pc, the odds of pc publishing a source in ac are roughly doubled.



# Preparing the dataset

Import required modules for renaming and world bank data:

```{r}
options(width=120);
require('plyr');
require('WDI');
require('MASS');
```

Read in raw file, calculate totals for each "other" (source / editor) country and "article" country.

```{r}
df <- read.table('../../../dat/source-counts-enhanced.tsv', header=TRUE, sep="\t", fileEncoding="utf-8");
totals <- rename(aggregate(count ~ project + article_country, df, sum), c("count"="article_total"));
df <- merge(df, totals, by=c("project","article_country"));
totals <- rename(aggregate(count ~ project + other_country, df, sum), c("count"="other_total"));
df <- merge(df, totals, by=c("project","other_country"));
totals <- rename(aggregate(count ~ project, df, sum), c("count"="project_total"));
df <- merge(df, totals, by=c("project"));
```

Read in World bank data and fields: GDP, Population, tertiary grads, broadband population.
Rename these fields and merge them into our table and impute zeros for missing data (TODO: something better).

```{r}
fields <- c("NY.GDP.MKTP.CD", "SP.POP.TOTL", "SE.TER.GRAD", "IT.NET.BBND");
field_mapping <- c("GDP", "POP", "EDU", "INTERNET");
names(field_mapping) = fields;

wdi = WDI(indicator=fields, start=2012, end=2012);
wdi$iso2c <- tolower(wdi$iso2c);
wdi <- rename(wdi, field_mapping);
df <- merge(df, wdi, by.x="other_country", by.y="iso2c");

df$GDP[is.na(df$GDP)] <- 0;
df$POP[is.na(df$POP)] <- 0;
df$EDU[is.na(df$EDU)] <- 0;
df$INTERNET[is.na(df$INTERNET)] <- 0;
```
Create log-transformed versions of skewed fields:

```{r}
df$logmigration1 <- log(df$migration1 + 2);
df$logmigration2 <- log(df$migration2 + 2);
df$logjournals1 <- log(df$journals1 + 2);
df$logjournals2 <- log(df$journals2 + 2);
df$lognewspapers1 <- log(df$newspapers1 + 2);
df$lognewspapers2 <- log(df$newspapers2 + 2);
df$logcountrylangpop <- log(df$countrylangpop + 2);
df$logdist <- log(df$dist + 2);
df$logGDP <- log(df$GDP + 2);
df$logPOP <- log(df$POP + 2);
df$logEDU <- log(df$EDU + 2);
df$logINTERNET <- log(df$INTERNET + 2);
```

# Overview of dataset

Our dataset models the relationship between language edition and the dipole of article country and the country of either the editor or source publisher (our two lenses of provenance).
We interchangebly calls the second country (editor or publisher) other_country or country2 throughout the dataset.

Four fields contain the key information being aggregated together by this dataset:

 * project: WP language edition
 * article_country: the country containing the article
 * other_country: the country associated with an editor or publisher
 * count: the number of edits or published citations associated with the triple (project, article_country, other_country) in our dataset.
 
The remaining fields containing attributes that we will use to model the relationship among these four fields.
For brevity, we'll use ac to reference the article country, oc to reference the publisher or source country, and wplang to refer to the wikipedia language.

 * article-is-native: whether the WP edition language is "natively" spoken by the ac
 * other-is-native: whether the WP edition language is "natively" spoken by the editor or pc
 * distance: distance between ac and oc
 * migration1: the number of people who migrated from ac to oc in 2012 (UN data)
 * migration2: the reverse of migration1
 * newspapers1: the number of newspapers published in ac (UN data)
 * newspapers2: the number of newspapers published in oc (UN data)
 * journals1: the number of journal articles published in ac (from Scopus via SCImago)
 * journals2: the number of journal articles published in oc (from Scopus via SCImago)
 * wplangshare: the fraction of worldwide speakers of wplang that live in oc
 * countrylangshare: across all languages used in ac, the fraction of worldwide speakers that live in oc 
 * countrylangpop: across all languages used in ac, the number of people who speak the language in the oc
 * langsim: the cosine similarity between the languages spoken in ac and oc
 
 Note that we use "raw counts" throughout most attributes (not per-captia values).
 The justification behind this is that work output is more related to the raw number of people (and therefore potential editors / publishers).
 A country with 1000 people, but 90% broadband access is going to have a much smaller pool of total Wikipedia work than a country with 1M people and 10% broadband access.

```{r}
head(df);
```

### Correlation between features

Below you can see the correlation matrix between all the features.
The first table contains actual values, and the second contains log-transformed values.

Note that the basic demographic indicators (GDP, POP, etc) are all strongly correlated with each other (0.7 to 0.95). 
The other variables are largely uncorrelated with each other, with a few moderate exceptions (e.g. migration in both directions = 0.4, journal vs newspaper is 0.36, population and newspaper is 0.76, etc.)

Interestingly, although journals and GDP are extremely highly correlated (0.98), log transforming decreases this relationship significantly to 0.52. 
This is probably because GDP and journals show high correlation in extreme values (e.g. the US), but after log transformation these have a much smaller effect. 
This same effect occurs for a variety of other values (e.g. INTERNET vs education).

Other variables show the opposite trend. 
For example, log transforming newspapers2 vs journals2 raises the correlation from 0.38 to 0.74.
This occurs when the two variables disagree on extreme values, but agree well on others.

```{r}
cor(unique(df[,c("distance", "GDP", "POP", "EDU", "INTERNET", "migration1", "migration2", "journals1", "journals2", "newspapers1", "newspapers2", "countrylangpop", "wplangshare", "langsim", "countrylangshare")]));

cor(unique(df[,c("logdist", "logGDP", "logPOP", "logEDU", "logINTERNET", "logmigration1", "logmigration2", "logjournals1", "logjournals2", "lognewspapers1", "lognewspapers2", "logcountrylangpop")]));
```

# Localness

First, we consider a statistical model of localness. We begin by filtering the dataset down to those records that have the same article-country and other-country:

```{r}
local_df <- subset(df, as.character(article_country) == as.character(other_country));
head(local_df);
nrow(local_df);
```

We now build the full model. 
Since article_country and other_country are the same, we only include variables for one of the countries.

```{r}
fit <- glm(count / article_total ~ article_is_native + newspapers1 + journals1 + wplangshare + GDP + EDU + INTERNET + lognewspapers1 + logjournals1 + logGDP + logPOP + logEDU + logINTERNET, family = binomial, data=local_df)
summary(fit);
```

We use R's step function to consider every possible model, adding one variable at a time.

We begin by carefully analyzing the effects of all variables individually. 
log(journals) is the strongest independent signal of localness, explaining 36% of the model deviance, followed by log(GDP).

```{r}
min_fit <- glm(count / article_total ~ 1, family = binomial, data=local_df);
s <- step(min_fit, scope=formula(fit), direction='forward', steps=1);
```

We now build up the full model in stepwise fashion (starting with log journals).
Since log(journals) is highly correlated with logGDP (0.52), it unsurprisingly pushes GDP to the bottom of the equation.

```{r}
s <- step(min_fit, scope=formula(fit), direction='forward', trace=FALSE);
s$anova;
```

We can now see that a few key variables can account for most of the model's explanatory power.
In particular, the first four variables: logjournals1, logPOP, article_is_native, and wplangshare explain about half of the deviance in the model.


```{r}
small_fit <- glm(count/ article_total ~ logjournals1 + logPOP + wplangshare, family = binomial, data=local_df)
summary(small_fit);
```

To interpret our model, we can look at the confidence intervals and odds ratios associated with it.

  * For every doubling in the number of journal articles in a country, the odds of a publisher / editor being local increase 105%
  * For every doubling in the the population of a country, the odds of a publisher / editor being local DECREASES by 61% (due to a colinearity with num journal articles)
  * If the country natively speaks the WP language edition, the oddsof a publisher / editor being local increases by 47%
  * For every increase of 10% in the amount a country represents of a language's worldwide base, the oddsof a publisher / editor being local increases by 6.8%. For example, for Japan (who represents almost all Japanese speakers) this would suggest an increase in odds of close to 70% ON TOP of the 47% for language edition.

```{r}
exp(cbind(oddsRatio=coef(small_fit), confint(small_fit)));
```



# Full dipole provenance

We now move on to modeling the dipole relationship.
For example, focusing on Japane, can how can we model how likely China is to be a publisher of a source?

### Modeling <language, country of source publisher>

We begin  by modeling top-level effects at the (language, source publisher country) level.
In other words, how likely is China to be the publisher of any citation in English Wikipedia?

We restrict our dataset to this top-level information:

```{r}
top_df <- unique(df[,c("project", "project_total", "other_total", "other_country", "other_is_native",  "newspapers2", "journals2", "wplangshare", "GDP", "POP", "EDU", "INTERNET", "lognewspapers2", "logjournals2", "logGDP", "logPOP", "logEDU", "logINTERNET")]);
head(top_df);
```

Now let's build and inspect a model.
Note that the "localness" problem appears to be much more difficult; this model explains 85% of the deviance in the dataset.

```{r}  
fit <- glm(other_total / project_total ~ . - other_country - project - project_total - other_total, family = binomial, data=top_df);
summary(fit);
```
Like before we will examine whether we can build a more compact and interpretable model.
We first look at the strength of each individual predictor.
Again, logJournals rises to the top.

```{r}
min_fit <- glm(other_total / project_total ~ 1, family = binomial, data=top_df);
s <- step(min_fit, scope=formula(fit), direction='forward', steps=1);
```

We now run a forward stepwise regression. 
Two variables, log journals and wplangshare rise to the top.

```{r}
s <- step(min_fit, scope=formula(fit), direction='forward', trace=FALSE);
s$anova;
```

We inspect the pruend model:

```{r}
small_fit <- glm(other_total/ project_total ~ logjournals2 + wplangshare, family = binomial, data=top_df);
summary(small_fit);
```

To interpret our model, we can look at the confidence intervals and odds ratios associated with it.

  * For every doubling in the number of journal articles in a country, the odds of a citation being from the country increase by 300%.
  * For every increase of 10% in the amount a country represents of a language's worldwide base, the odds a citation coming froms that country increase by 600%.

```{r}
exp(cbind(oddsRatio=coef(small_fit), confint(small_fit)));
```


### Modeling dipoles

Finally, we move on to dipoles. 
Since we have already accounted for top-level effects, we introduce a variable that models the fraction of all citations a country accounts for in each language edition.

```{r}
df$top_level <- df$other_total / df$project_total;
df$is_local <- (as.character(df$article_country) == as.character(df$other_country));
#df$article_country;
#df$other_country;
head(df);
```

We now try to model the remaining deviance (top-level explains 22% of variance):

```{r}
summary(glm(count / article_total ~ top_level, family = binomial, data=df));
```

Once again, all terms in the model are extremely significant and the full model accounts for the vast majority (92%) of the variance.

```{r}

full <- glm(count / article_total ~ top_level + countrylangshare + countrylangpop + langsim + migration1 + migration2 + distance + logdist+ logmigration1 + logmigration2 + is_local, family = binomial, data=df);
summary(full);
```

We now try to identify a smaller, more interpretable model.
Unsurprisingly, the background popularity (analyzed in the previous section) of a country as a source publisher is by far the strongest term.

We then move on to the second term in the model.
log(distance) is most important, followed by countrylangshare.

```{r}
min_fit <- glm(count/ article_total ~ 1, family = binomial, data=df);
s <- step(min_fit, scope=formula(full), direction='forward', steps=2);
s <- step(min_fit, scope=formula(full), direction='forward', steps=6, trace=FALSE);
s$anova;
```

```{r}
small_fit <- glm(count / article_total ~ top_level + is_local +  logmigration1, family = binomial(logit), data=df);
summary(small_fit);
```

To interpret our model, we can look at the confidence intervals and odds ratios associated with it.

  * For every doubling in the distance between countries, the odds of a country being associated an article in another country are roughly halved.
  * For every doubling in the number of journal articles for a country, the odds of a publisher in the country being cited roughly double.

```{r}
exp(cbind(oddsRatio=coef(small_fit), confint(small_fit)));
```
